Introducing statistics with \\texttt{intRo}

### 2014-12-04 student-paper-competition-2014/paper.Rmd
small typos

---
title: "Introducing statistics with \\texttt{intRo}"
author: "Andee Kaplan & Eric Hare"
date: "December 14, 2014"
output:
  pdf_document:
    includes:
      in_header: preamble.tex
    number_sections: yes
bibliography: refs.bib
---

\begin{abstract}
\texttt{intRo} is a modern web-based application for performing basic data analysis and statistical routines. Leveraging the power of \texttt{R} and \texttt{Shiny}, \texttt{intRo} implements common statistical functions in a powerful and extensible modular structure, while remaining simple enough for the novice statistician. This simplicity lends itself to a natural presentation in an introductory statistics course as a substitute for other commonly used statistical software packages, such as \texttt{Excel} and \texttt{JMP}. \texttt{intRo} is currently deployed at the URL \url{http://www.intro-stats.com}
\end{abstract}

@@

# Introduction
The widespread adoption of \texttt{R} as a tool for statistical analysis has undoubtedly been a great development for the scientific community. However, using \texttt{R} in most cases still requires a basic knowledge of programming concepts which may be unfamiliar to the introductory statistics student. This issue may explain why introductory courses often utilize software applications such as \texttt{JMP}, even though instructors would tend to use a tool like \texttt{R} in their own work. Still, signficant compromises must be made when using software like \texttt{JM}P, including dealing with software licenses, unsupported desktop platforms, and non-ideal user interface decisions. In teaching STAT 226 at Iowa State University, we witnessed profound struggles by students attempting to practice the introductory concepts discussed in class. Often, these struggles were directly related to the use of the software and not any sort of fundamental misunderstanding of the material.

@@

Within the data category there are two modules, sources and transformation. The data sources module allows the user to select or upload a dataset. Several datasets are pre-included, which can be selected via a drop-down menu. The user can check a box to upload a dataset, and then select the CSV file from his computer. Upon selection or successful upload of a dataset, a table view of the data will be visible on the right. From this module the user can select a subset of the data either by value or random selection. The dataset selected in the data module will be made available to all other modules of intRo. The data transformation module allows transformation of numeric variables to categorical and categorical variable to numeric. Additionally, numeric variable can be transformed using a power transformation. The transformed variables can be saved to the dataset for use elsewhere in \texttt{intRo}.

@@

Despite the current functionality, there are some practical improvements in the works that will make \texttt{intRo} more accessible and useful. In particular, we have begun development on an \texttt{R} package which will allow modules to be created automatically. This package will generate the necessary file structure to allow the module's incorporation into \texttt{intRo}. This will vastly improve intRo's flexibility and allow it to be used in a wider range of curriculums, including more advanced statistics courses. We will also be trialing \texttt{intRo}'s use in a statistics course at Iowa State University in Spring Semester 2015. This will allow us to assess its usefulness relative to software used in the past, as well as areas for improvement. Challenges exist with regards to the wider adoption of \texttt{intRo}. For instance, we will need to monitor how well the server \texttt{intRo} is hosted on handles the load of dozens of students performing data analyses. If performance issues are noted, the infrastructure used may need to be expanded to handle current and future load.

>>>>

---
title: "Introducing statistics with \\texttt{intRo}"
author: "Eric Hare & Andee Kaplan"
date: "December 14, 2014"
output:
  pdf_document:
    includes:
      in_header: preamble.tex
    number_sections: yes
bibliography: refs.bib
---

\begin{abstract}
\texttt{intRo} is a modern web-based application for performing basic data analysis and statistical routines. Leveraging the power of \texttt{R} and \texttt{Shiny}, \texttt{intRo} implements common statistical functions in a powerful and extensible modular structure, while remaining simple enough for the novice statistician. This simplicity lends itself to a natural presentation in an introductory statistics course as a substitute for other commonly used statistical software packages, such as \texttt{Excel} and \texttt{JMP}. \texttt{intRo} is currently deployed at the URL \url{http://www.intro-stats.com}. Within this paper we introduce the application and explore the design decisions underlying \texttt{intRo} as well as highlight some challenges and advantages of reactive programming.
\end{abstract}

@@

# Introduction
The widespread adoption of \texttt{R} as a tool for statistical analysis has undoubtedly been a great development for the scientific community. However, using \texttt{R} in most cases still requires a basic knowledge of programming concepts which may be unfamiliar to the introductory statistics student. This issue may explain why introductory courses often utilize software applications such as \texttt{JMP}, even though instructors would tend to use a tool like \texttt{R} in their own work. Still, signficant compromises must be made when using software like \texttt{JMP}, including dealing with software licenses, unsupported desktop platforms, and non-ideal user interface decisions. In teaching STAT 226 at Iowa State University, we witnessed profound struggles by students attempting to practice the introductory concepts discussed in class. Often, these struggles were directly related to the use of the software and not any sort of fundamental misunderstanding of the material.

@@

Within the data category there are two modules, sources and transformation. The data sources module allows the user to select or upload a dataset. Several datasets are pre-included, which can be selected via a drop-down menu. The user can check a box to upload a dataset, and then select the CSV file from his computer. Upon selection or successful upload of a dataset, a table view of the data will be visible on the right. From this module the user can select a subset of the data either by value or random selection. The dataset selected in the data module will be made available to all other modules of \texttt{intRo}. The data transformation module allows transformation of numeric variables to categorical and categorical variable to numeric. Additionally, numeric variable can be transformed using a power transformation. The transformed variables can be saved to the dataset for use elsewhere in \texttt{intRo}.

@@

Despite the current functionality, there are some practical improvements in the works that will make \texttt{intRo} more accessible and useful. In particular, we have begun development on an \texttt{R} package which will allow modules to be created automatically. This package will generate the necessary file structure to allow the module's incorporation into \texttt{intRo}. This will vastly improve \texttt{intRo}'s flexibility and allow it to be used in a wider range of curriculums, including more advanced statistics courses. We will also be trialing \texttt{intRo}'s use in a statistics course at Iowa State University in Spring Semester 2015. This will allow us to assess its usefulness relative to software used in the past, as well as areas for improvement. Challenges exist with regards to the wider adoption of \texttt{intRo}. For instance, we will need to monitor how well the server \texttt{intRo} is hosted on handles the load of dozens of students performing data analyses. If performance issues are noted, the infrastructure used may need to be expanded to handle current and future load.
